<!--- Time-stamp: <2013-11-27 11:56:25 chl> -->

<!--- To generate HTML output:
library(knitr)
library(markdown)
knit("lattice.rmd", quiet=TRUE)
markdownToHTML("lattice.md", "lattice.html", stylesheet="styles.css", option=c("highlight_code", "toc"), title="Using lattice graphics")
browseURL("lattice.html")
-->


```{r setup, include=FALSE}
opts_knit$set(progress=FALSE, verbose=FALSE, width=90)
opts_chunk$set(message=FALSE, tidy=TRUE, highlight=TRUE, fig.align="center")
library(gridExtra)
library(latticeExtra)
trellis.par.set(custom.theme.2())
trellis.par.set(plot.symbol=list(pch=19, cex=1.2),
                strip.shingle = list(col = "darkgoldenrod1"),
                strip.background = list(col = "transparent"), 
                fontsize = list(text = 16, points = 8))
```

<p style="font-size: 200%; font-weight: bold; text-align: center;">Using lattice graphics</p>


This document describes the main features of R's [lattice][1] package for
use as a graphical exploratory tool or produce ready-to-print graphical
figures. Extra features are available in the [latticeExtra][2] packages, and
more information can be found on the [latticeExtra companion website][3], as
well as the [lattice companion website][4]. The definitive reference to
using lattice graphics is

> Sarkar, D. (2007). *Lattice: Multivariate Data Visualization with
> R*. Springer.


## Introduction

A definitive key feature of `lattice` is that it will generally
produce informative and already nice looking plots. Furthermore, as it
relies on "formula" notation, calling a `lattice` plotting function
is not so much different from evaluating statistical models using
`lm()` or `glm()`. Finally, the updating mechanism renders
step-by-step plotting interaction an easy task to perform, even for
users with limited knowledge of R; in fact, there's no real need to
bother with redrawing onto the same graphical device using `points()`
or `lines()` instead of `plot()`: with `lattice`, just draw a basic
canvas, and then update it with additional information or replace
default settings (e.g., color, type of points) with customized ones.

For R practitioners used to base graphics, switching to `lattice`
will not be a big deal. The base syntax for drawing a simple
scatterplot is `plot(x, y)` or `plot(y ~ x, data=)` which reads
`xyplot(y ~ x, data=)` using `lattice`. Most of the base plotting
commands have their equivalent in `lattice`, for example:

| Type        | Base R            | Lattice              |
| ----------- |:-----------------:|:--------------------:|
| Scatterplot | plot(x,y)         | xyplot(y ~ x)        |
| Histogram   | hist(x)           | histogram(~ x)       |
| Density     | plot(density(x))  | densityplot(~ x)     |
| Barchart    | barplot(table(x)) | barchart(xtabs(~ x)) |
| Dotchart    | dotchart(table(x))| dotplot(x)           |
| QQ plot     | qqplot(x)         | qqmath(~ x)          |


However, where we would have to split our plot in multiple subplot manually
using base R graphics, `lattice` offers a convenient and intuitive way for
arranging the information into multiple panels using the
[trellis approach][5] used and described in Cleveland's famous
books:

- Cleveland, WS, [The Elements of Graphing Data](http://goo.gl/NmC9to), Hobart Press, 1994
- Cleveland, WS, [Visualizing Data](http://goo.gl/mmnt6k), Hobart Press, 1993

Another good reference, which put more emphasis on R implementation
is Murrell's [*R Graphics*][6].


## Main components of a lattice object

The main components of a `lattice` object are shown in the picture
below. In this case, an automatic layout was choosen where each panel
describes the relationship between the outcome, `y`, and one
predictor, `x`, according to the levels of a categorical variable `g`
and a continuous variable `z` summarized as *shingles*.

Let's start by drawing a basic scatterplot for a multiway data set. Our data
set is composed of four variables that are simulated very basically as shown
below. Variables `x` and `z` are simply drawn from an uniform distribution,
while `g` is a random sample of `a` and `b`. Then, `z` is discretized using
the `equal.count()` function, which basically amounts to dividing z-values
in distinct classes. Contrary to `cut()`, it allows partial overlapping
between those classes, meaning that an observation belong to several classes
at once. This is the basis of `lattice`'s shingles, and it permit the
cross-classification of several variables, including continuous ones.
```{r fig1}
set.seed(88)
n <- 100
x <- runif(n, min=10, max=20)
z <- runif(n, min=0, max=10)
g <- factor(sample(letters[1:2], n, rep=TRUE))
y <- 1.2 + .8*x - .2*z + .5*as.numeric(g) + rnorm(n)

d <- data.frame(x, z, g, y)
d <- transform(d, zc = equal.count(z, 4))

xyplot(y ~ x | g + zc, data=d,
       type=c("p", "g", "smooth"), col.line="peachpuff3", lwd=2,
       xlab="X-axis label", ylab="Y-axis label", main="Title", sub="Subtitle")
```

It is worth noting that a scatterplot smoother (lowess) is applied to all 3
panels defined by the relation `x` conditional to `g` and `zc`. All axes are
scaled in a similar manner, and horizontal and vertical axis alternate from
bottom to top, or left to right, respectively. Labels for the x- and y-axes
are controlled as in base graphics, using `xlab=` and `ylab=`, and it is
also possible to directly use `par` parameters like `lwd=` (line width). The
`col.line=` option is specific to the `panel.smooth()` function, which is
used internally as soon as we add a `type="smooth"` option. The formula used
above follows R's convention, except that `*` which denotes interaction
between variables is meaningless (in most cases).

Other important parameters of a `lattice` display are given below:

- `aspect` determine the ratio of height to width of the graphical
  region. It can be a (positive) number, or one of the available options:
  `xy` means that the *45° banking rule* [@cleveland88] is used to
  determine x and y scale; `iso` means that the number of units per cm
  is the same for both axes which implies that relation between physical
  distance on the display and in the data scale is respected.
- `layout` describes the arrangement of panel as columns by rows;
  e.g., `layout=c(2, 3)` means to display the panel in a grid composed
  of two columns and three rows. 
- `scales` is a list with various parameters to customize axis setting.
- `between` is the amount of space between each panel. By default, it
  equals 0, that is there is no space left between panels.
- `key` or `auto.key` allows to set up the legend, with more or less
  default control.


As an example, consider the following variations around the scatterplot.


```{r fig2}
library(hexbin)
n <- 1000
d <- transform(data.frame(x=runif(n, 0, 10)),
               y=0.7*x + rnorm(n))
d <- d[sample(1:n, n, replace=TRUE),]

p <- list()
p[[1]] <- xyplot(y ~ x, data=d, col="cornflowerblue")
p[[2]] <- update(p[[1]], pch=19, col=rgb(0, 0, .8, .25),
                 cex=.6, sub="col=rgb(0, 0, .8, .25)")
p[[3]] <- update(p[[2]], panel=panel.hexbinplot,
                 sub="panel=panel.hexbinplot")
p[[4]] <- update(p[[1]], jitter.data=TRUE, pch="+",
                 type=c("p", "g", "smooth"),
                 col.line="grey20", lwd=2,
                 sub="jitter.data=TRUE")
do.call(grid.arrange, p)  ## require to load the gridExtra package
```

In this simulated dataset, we have introduced around one third of replicates
to show how easy it is to cope with this situation using transparency
(`alpha=` or `col=`), hexagonal binning from the [hexbin][7] package, or
jittering (`jitter.data=`). What is important here is to realize that the
above customizations do not use anything else than specific arguments to
`xyplot()`. That is to say we often won't need to write complex R code to
get the desired effect. However, the updating method used in the above code
alleviates the need to write the same command several times when we just
want to change a given parameter or the overall layout.

But, look at the complete definition of the `xyplot()` function:

     xyplot(x,
            data,
            allow.multiple = is.null(groups) || outer,
            outer = !is.null(groups),
            auto.key = FALSE,
            aspect = "fill",
            panel = lattice.getOption("panel.xyplot"),
            prepanel = NULL,
            scales = list(),
            strip = TRUE,
            groups = NULL,
            xlab,
            xlim,
            ylab,
            ylim,
            drop.unused.levels = lattice.getOption("drop.unused.levels"),
            ...,
            lattice.options = NULL,
            default.scales,
            default.prepanel = lattice.getOption("prepanel.default.xyplot"),
            subscripts = !is.null(groups),
            subset = TRUE)
     
Some formal parameters are `NULL` by default (e.g., `auto.key=`, `scales=`),
but more interestingly there is the *special variable length argument*
(`...`) that appears among the various options. It is important to
understand that it is used to pass arguments to the `panel()` function,
other than default options. For example, to change the color a smoothing
line invoked with `type="smooth"`, we would add `col.line=` to the main
`xyplot()` call. E.g.,

```{r fig3, eval=FALSE}
xyplot(y ~ x, type=c("p", "smooth"), col.line="red", lwd=2)
```
will draw a local smoother in red where line width (`lwd=`) is two times the
default line width.

## Customizing default plots

Sometimes, it may be useful to add contextual information to an existing
graphic. Consider the following example where pre and post measurements on
weight have been made for three groups of young female anorexia patients:
```{r fig4}
data(anorexia, package="MASS")
p <- xyplot(Postwt ~ Prewt | Treat, data=anorexia,
            layout=c(3,1), aspect="iso")
p
```

Since measurements are expressed on the same scale, we added the
`aspect="iso"` option, which ensures that teh graphical display will reflect
the same variation on the x- and y-axis.

A scatterplot is appropriate when individual measurements are paired such
that weight increase will be reflected by a cloud of points located above
the identity line. However, this one is lacking which makes it difficult to
get a reliable sense of the tendency in this plot. To add a 45° line, we
could simply update the above plot and add another panel like this:
```{r fig5}
p <- update(p, abline=list(a=0, b=1, col="lightgrey"))
print(p)
```

Now, it is evident that patients under family treatment (FT) have seen their
weight increased more consistently and by a larger amount,compared to other
patients. There is a `print` method for lattice graphic. It is not necessary
to use it to display a plot on the active display, but it comes with
additional parameters


[1]: http://cran.r-project.org/web/packages/lattice/index.html
[2]: http://cran.r-project.org/web/packages/latticeExtra/index.html
[3]: http://latticeextra.r-forge.r-project.org/
[4]: http://lmdvr.r-forge.r-project.org/
[5]: http://stat.bell-labs.com/project/trellis/wwww.html
[6]: https://www.stat.auckland.ac.nz/~paul/RGraphics/rgraphics.html
[7]: http://cran.r-project.org/web/packages/hexbin/index.html
